---
title: "SuperLearner TI-IDF SVD"
author: "Heather, Jarrett, Andras"
output: html_document
---

## SuperLearner Ensemble

### 1. Overview

Base learners used:

- Elastic-net logistic regression (very strong baseline)
- Linear SVM
- XGBoost (low depth)
- Random forest (optional but weaker here)
- Naive Bayes (usually weak after SVD -- often drop)

### 2. Install packages and load data

```{r install, message=FALSE, warning=FALSE}
#install.packages(c(
#  "SuperLearner",
#  "glmnet",
#  "e1071",
#  "xgboost",
#  "randomForest",
#  "data.table",
#  "PRROC",
#  "pROC"
#))

```

```{r libraries, message=FALSE, warning=FALSE}
# Load all required libraries after installation
library(tidyverse)
library(SuperLearner)
library(xgboost)
library(glmnet) # Explicitly load glmnet
library(e1071)  # For SL.svm (even if commented out in learners, good practice)
library(randomForest)
library(data.table)
library(PRROC)
library(pROC)
library(parallel)
library(doParallel)
library(here)

```

```{r load-data}
# Load data
dt_tfidf <- readRDS(here('04_output', "tfidf_svd_features.rds"))
dt_doc2vec <- readRDS(here('04_output', 'doc2vec_features.rds' ))
dt_pubmed <- readRDS(here('04_output', 'pubmedbert_embeddings.rds'))

cv_folds <- readRDS(here('04_output',"cv_folds.rds"))

```

### 3. Define outcome & predictors

```{r outcome-predictors}
# Separate the response from features
Y <- dt_tfidf[[2]]
X <- dt_tfidf[, -c(1, 2)]
```

```{r coerce-binary}
# Check that the response is binary
Y <- as.numeric(Y)
```

### 4. Compute class weights (almost always necessary)

```{r class-weights}
# Compute class weights
pos_rate <- mean(Y == 1)
neg_rate <- mean(Y == 0)

# Assign inverse class frequency weights to balance the classes
obs_weights <- ifelse(Y == 1, 1/pos_rate, 1/neg_rate)

# This is crucial for screening recall.
```

```{r custom_svm}
#use a c-svm instead of the defult nu-svm for better performance on imbalanced data. Also set class weights to handle imbalance
SL.svm.custom <- function(Y, X, newX, family, obsWeights, ...) {
  library(e1071)
  fit <- svm(
    x = X, 
    y = as.factor(Y),
    type = "C-classification",  # C-SVM instead of nu-SVM
    kernel = "radial",
    class.weights = c("0" = 1, "1" = sum(Y == 0) / sum(Y == 1)),  # handle imbalance
    probability = TRUE
  )
  pred <- predict(fit, newdata = newX, probability = TRUE)
  pred <- attr(pred, "probabilities")[, "1"]
  return(list(pred = pred, fit = fit))
}
```

### 5. Define TF-IDF-optimized SuperLearner library

```{r learners}
# Set learners' library
learners <- c(
  "SL.glmnet",        # Elastic-net logistic (BEST for TF-IDF)
  #"SL.svm",           # Linear SVM
  "SL.xgboost",       # Shallow boosted trees
  "SL.randomForest",    # Optional, usually lower weight
  "SL.naiveBayes",     # Naive Bayes
  "SL.svm.custom" # Custom SVM with class weights for imbalance 
)
```

### 6. Run 5-fold SuperLearner cross-validation

```{r cv-superlearner}

cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
clusterSetRNGStream(cl, iseed = 42)
clusterExport(cl, list(
  "Y", "X", "learners", "obs_weights", "cv_folds"
))

# Run cross validation
set.seed(542)

sl_cv <- CV.SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),
  obsWeights = obs_weights,
  method = "method.NNLS",
  cvControl = list(V = 5, validRows = cv_folds), # when passing the folds set V here to the number of folds expected
  parallel = cl
)

stopCluster(cl)

# Produces out-of-fold predictions. Prevents leakage. Suitable for reporting
```

### 7. Inspect ensemble weights

```{r ensemble-weights}
# Print weights
sl_cv$SL.weights

# Typical TF-IDF outcome: SL.glmnet: dominant. SL.svm: meaningful.
# SL.xgboost: smaller but helpful. SL.randomForest: often near zero
# This is expected and correct.
```

### 8. Evaluate PR-AUC (preferred for screening)

```{r pr-auc}
# Sketch PR Curve
library(PRROC)

pr <- pr.curve(
  scores.class0 = sl_cv$SL.predict[Y == 1],
  scores.class1 = sl_cv$SL.predict[Y == 0],
  curve = TRUE
)

pr$auc.integral
```

### 9. Recall-constrained threshold selection (screening-ready)

```{r threshold-selection}
# Search for recall selection (>=95% recall)
target_recall <- 0.95
threshold_grid <- seq(0.01, 0.6, by = 0.005)

recalls <- sapply(threshold_grid, function(t) {
  pred <- ifelse(sl_cv$SL.predict >= t, 1, 0)
  sum(pred == 1 & Y == 1) / sum(Y == 1)
})

best_threshold <- min(threshold_grid[recalls >= target_recall])
best_threshold
```

```{r workload-reduction}
# Workload reduction at that threshold (important result)
pred_final <- ifelse(sl_cv$SL.predict >= best_threshold, 1, 0)
workload_reduction <- 1 - mean(pred_final == 1)

workload_reduction
```

### 10. False negatives (must be audited)

```{r false-negatives}
# Check false negatives
false_negatives <- which(Y == 1 & pred_final == 0)
length(false_negatives)
```

### 11. Final SuperLearner model (for deployment only)

```{r final-model}
# Set predict model
final_sl <- SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),
  obsWeights = obs_weights,
  method = "method.NNLS"
)
```

### 12. Prediction pipeline

```{r prediction-pipeline}
new_dt <- readRDS("tfidf_svd_features_new.rds")
X_new <- new_dt[, -c(1, 2), with = FALSE]

pred_prob <- predict(
  final_sl,
  newdata = X_new,
  onlySL = TRUE
)$pred

pred_label <- ifelse(pred_prob >= best_threshold, 1, 0)

output <- data.table(
  article_id = new_dt[[1]],
  prob_include = pred_prob,
  screen_in = pred_label
)
```
