---
title: "Text Preparation"
format: html
editor: visual
editor_options:
  chunk_output_type: console
execute: 
    eval: false
---

```{r setup}
library(tidyverse)
library(caret)
library(quanteda)
library(here)
library(reticulate)
use_condaenv('r-gensim', required = TRUE)
```

# Text Preparation: PubMedBERT vs TF-IDF vs Doc2Vec

## 1: Set up for 5-Fold Cross Validation

Create stratified 5-fold CV splits. The `folds` list is saved as an `.rds` file so the same fold assignments can be reused in `model_comparison.qmd`.

```{r load_cleaned_abstracts}
# Load cleaned abstracts and labels; assign a row ID to each record
# str_squish() collapses embedded \n, \r\n, \t to single spaces so that
# the text field is single-line and safe for CSV/Python parsing
abstracts <- read_csv(here('04_output', 'test_data.csv')) %>%
    rowid_to_column('id') %>%
    select(id, text, recd_screen) %>%
    mutate(text = str_squish(text))

# Set seed for reproducibility
set.seed(42)

# Create stratified 5-fold CV splits
# folds[[k]] contains the row indices (= id values) for the test set of fold k
folds <- createFolds(abstracts$recd_screen, k = 5, list = TRUE, returnTrain = FALSE)

# Check stratification across folds
map_dfr(seq_along(folds), ~ tibble(
  fold     = .x,
  n        = length(folds[[.x]]),
  pos_rate = mean(abstracts$recd_screen[folds[[.x]]])
))

# Save folds object for use in model_comparison.qmd
saveRDS(folds, here('04_output', 'cv_folds.rds'))

# Save abstracts for Python embedding generation
write_csv(abstracts, here('04_output', 'abstracts.csv'))
```

------------------------------------------------------------------------

# Text Pre-processing Methods

## Method 1: PubMedBERT

Embeddings are generated in Python using the pre-trained `NeuML/pubmedbert-base-embeddings` model. Since the model is pre-trained, embeddings are fold-independent — all records are encoded at once. Fold awareness is handled in R via `cv_folds.rds`.

Run in Google Colab with GPU runtime for faster embedding generation.

```{python}
import pandas as pd
from sentence_transformers import SentenceTransformer

# Load abstracts (id, text, recd_screen)
df = pd.read_csv('abstracts.csv')

# Encode all records at once
model = SentenceTransformer('NeuML/pubmedbert-base-embeddings')
embeddings = model.encode(df['text'].tolist(), batch_size=64, show_progress_bar=True)

emb_df = pd.DataFrame(embeddings)
emb_df.insert(0, 'id', df['id'].values)
emb_df.insert(1, 'recd_screen', df['recd_screen'].values)

emb_df.to_csv('pubmedbert_embeddings.csv', index=False)
print(f"Final embeddings shape: {emb_df.shape}")
```

### Load PubMedBERT embeddings into R

```{r load_bert}
# Read back the full embeddings matrix with label column
pubmedbert <- read_csv(here('04_output', 'pubmedbert_embeddings.csv'))
```

------------------------------------------------------------------------

## Method 2: TF-IDF

TF-IDF features are computed in R. To prevent data leakage, the vocabulary is learned from the training folds only, and `dfm_match()` aligns the test fold feature set to the training vocabulary. Test-fold feature matrices are assembled into a single full-dataset matrix after the fold loop.

-   Lowercase
-   Remove punctuation, numbers, symbols
-   Remove stopwords (Snowball English list)
-   Stem words (Porter algorithm)

```{r tfidf_function}
# TF-IDF preprocessing function
preprocess_tfidf <- function(texts) {
  tokens(texts,
         remove_punct   = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) %>%
    tokens_tolower() %>%
    tokens_remove(pattern = stopwords("english", source = "snowball")) %>%
    tokens_wordstem(language = "english") %>%
    dfm()
}
```

```{r tfidf_vocab_diagnostic}
# Build DFM on full corpus to explore vocabulary size vs. min_docfreq threshold
# Use this plot to choose min_docfreq in the fold loop below
dfm_full <- preprocess_tfidf(abstracts$text)

vocab_diag <- map_dfr(c(1, 2, 3, 5, 10, 15, 20, 30, 50, 75, 100), ~ tibble(
  min_docfreq  = .x,
  n_terms      = nfeat(dfm_trim(dfm_full, min_docfreq = .x)),
  pct_retained = nfeat(dfm_trim(dfm_full, min_docfreq = .x)) / nfeat(dfm_full) * 100
))

ggplot(vocab_diag, aes(x = min_docfreq, y = n_terms)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(1, 2, 3, 5, 10, 15, 20, 30, 50, 75, 100)) +
  labs(title = "Vocabulary size vs. minimum document frequency",
       x     = "min_docfreq (minimum documents a term must appear in)",
       y     = "Number of terms retained") +
  theme_minimal()

print(vocab_diag)
```

```{r tfidf_dropped_terms}
# Inspect terms that would be dropped at a given threshold
# Adjust `inspect_threshold` to match whichever cutoff you are considering
inspect_threshold <- 10

doc_freqs <- docfreq(dfm_full)

tibble(term    = names(doc_freqs),
       docfreq = as.integer(doc_freqs)) %>%
  filter(docfreq < inspect_threshold) %>%
  arrange(desc(docfreq)) %>%   # highest docfreq first = most borderline drops
  print(n = 50)
```

```{r tfidf_folds}
# List to collect test-fold feature matrices in row order
tfidf_test_list <- vector("list", 5)

for (fold_index in 1:5) {
  cat(sprintf("\n== TF-IDF Fold %d ==\n", fold_index))

  test_idx  <- folds[[fold_index]]
  train_idx <- setdiff(seq_len(nrow(abstracts)), test_idx)

  df_train <- abstracts[train_idx, ]
  df_test  <- abstracts[test_idx, ]

  # Learn vocabulary from training fold; trim rare terms; apply to test fold
  # min_docfreq = 5 drops terms appearing in fewer than 5 training documents,
  # keeping vocabulary manageable and avoiding sparse/noisy features
  dfm_train <- preprocess_tfidf(df_train$text) %>%
      dfm_trim(min_docfreq = 5)
  dfm_test  <- preprocess_tfidf(df_test$text) %>%
      dfm_match(features = featnames(dfm_train))

  # Apply TF-IDF weighting using training-corpus IDF (prevents leakage)
  idf_train      <- docfreq(dfm_train, scheme = "inverse", base = 10)
  dfm_test_tfidf <- dfm_weight(dfm_test, weights = idf_train)

  # Store test-fold matrix with metadata
  tfidf_test_list[[fold_index]] <- as.data.frame(as.matrix(dfm_test_tfidf)) %>%
      mutate(id          = df_test$id,
             recd_screen = df_test$recd_screen,
             .before     = 1)
}

# Assemble full-dataset TF-IDF matrix, sorted by id
tfidf_full <- bind_rows(tfidf_test_list) %>%
    arrange(id)

write_csv(tfidf_full, here('04_output', 'tfidf_features.csv'))
```

------------------------------------------------------------------------

## Method 3: Doc2Vec

Doc2Vec embeddings are computed in R using `paragraph2vec()`. To prevent data leakage, a separate model is trained on the training folds within each CV fold. Test-fold embeddings are assembled into a single full-dataset matrix after the fold loop.

-   Lowercase
-   Keep only alphanumeric characters and spaces

```{r doc2vec_function}
# Doc2Vec preprocessing: lowercase + alphanumeric only + normalize whitespace
preprocess_doc2vec <- function(.data) {
  .data %>%
    mutate(text = str_to_lower(text),
           text = str_replace_all(text, "[^a-z0-9]", " "),
           text = str_squish(text))
}
```

```{r doc2vec_setup}
# Import gensim Doc2Vec via reticulate
# Requires gensim in the active Python environment:
#   py_install("gensim")  # run once if not installed
TaggedDocument <- import("gensim.models.doc2vec")$TaggedDocument
Doc2Vec        <- import("gensim.models.doc2vec")$Doc2Vec

# Define a Python batch inference function to avoid crossing the R-Python
# boundary once per document (which causes C stack overflow at scale)
py_run_string("
def batch_infer(model, texts):
    # Return list of lists (not numpy array) -- reticulate converts reliably
    return [model.infer_vector(t.split()).tolist() for t in texts]
")
```

```{r doc2vec_folds}
# List to collect test-fold embeddings in row order
doc2vec_test_list <- vector("list", 5)

for (fold_index in 1:5) {
  cat(sprintf("\n== Doc2Vec Fold %d ==\n", fold_index))

  test_idx  <- folds[[fold_index]]
  train_idx <- setdiff(seq_len(nrow(abstracts)), test_idx)

  df_train <- abstracts[train_idx, ] %>% preprocess_doc2vec()
  df_test  <- abstracts[test_idx,  ] %>% preprocess_doc2vec()

  # Build tagged training documents for gensim
  # Each document: words = token list, tags = unique doc id string
  tagged_train <- lapply(seq_len(nrow(df_train)), function(i) {
    TaggedDocument(
      words = strsplit(df_train$text[i], " ")[[1]],
      tags  = list(as.character(df_train$id[i]))
    )
  })

  # Train Doc2Vec on training fold only
  model <- Doc2Vec(
    tagged_train,
    vector_size = 100L,
    window      = 5L,
    min_count   = 2L,
    epochs      = 5L,
    workers     = 4L,
    seed        = 42L
  )

  # Infer embeddings for test fold -- entire loop runs in Python to avoid
  # repeated R-Python boundary crossings that cause C stack overflow
  # batch_infer returns a list of 100-element lists; rbind assembles into matrix
  test_embeddings <- do.call(rbind, py$batch_infer(model, df_test$text))

  # Store with metadata -- row order matches df_test
  doc2vec_test_list[[fold_index]] <- as.data.frame(test_embeddings) %>%
      mutate(id          = df_test$id,
             recd_screen = df_test$recd_screen,
             .before     = 1)
}

# Assemble full-dataset Doc2Vec matrix, sorted by id
doc2vec_full <- bind_rows(doc2vec_test_list) %>%
    arrange(id)

write_csv(doc2vec_full, here('04_output', 'doc2vec_features.csv'))

```

------------------------------------------------------------------------

## Outputs

| File | Description |
|----|----|
| `04_output/abstracts.csv` | Full text data (id, text, recd_screen) for Python |
| `04_output/cv_folds.rds` | Fold assignments list (used by text prep + modeling) |
| `04_output/pubmedbert_embeddings.csv` | 768-dim PubMedBERT embeddings + label |
| `04_output/tfidf_features.csv` | TF-IDF feature matrix + label |
| `04_output/doc2vec_features.csv` | Doc2Vec embeddings (100-dim) + label |

------------------------------------------------------------------------

## Key Implementation Notes

### 1. **Data Leakage Prevention**

-   TF-IDF: Vocabulary and IDF learned from training folds only; `dfm_match()` aligns test vocabulary; `docfreq()` extracts training IDF applied via `dfm_weight()`
-   Doc2Vec: Model trained on training folds only; `infer_vector()` (gensim via reticulate) infers test embeddings with frozen word weights
-   PubMedBERT: Pre-trained — no leakage risk; all records encoded at once

### 2. **Preprocessing Differences**

-   **TF-IDF**: Aggressive (stop words, stemming) → needed for sparse representations
-   **Doc2Vec**: Light (lowercasing, alphanumeric only) → learns from context
-   **PubMedBERT**: None → pre-trained model handles this internally
