---
title: "SuperLearner PubMedBERT Embedding"
author: "Heather, Jarrett, Andras"
output: html_document
---

## Packages and Data

### 1. Install & load packages

```{r install, message=FALSE, warning=FALSE}
# Install packages
#install.packages(c(
#  "SuperLearner",
#  "glmnet",
#  "e1071",
#  "xgboost",
#  "randomForest",
#  "naivebayes",
#  "data.table"
#))

library(SuperLearner)
library(data.table)
```

```{r load-kernlab, message=FALSE, warning=FALSE}
# Load other libraries
library(kernlab)
```

```{r set-seed}
# Set seed to ensure reproducibility
set.seed(542)
```

### 2. Load PubMedBERT embeddings

```{r load-data}
# Load the data
data <- readRDS("pubmedbert_embeddings.rds")
```

```{r outcome-predictors}
# Prepare the response (Y) and predictors (X)
Y <- data[, 2]
X <- as.data.frame(data[, -2])
```

## Base Learners Individually

### Logistic Regression Base Learner

```{r lr-library}
# Define the base learner (Logistic Regression)
sl_library <- c("SL.glm")
```

```{r lr-cv}
# Run the Cross-Validated SuperLearner
set.seed(542) # For reproducibility
cv_sl <- CV.SuperLearner(
  Y = Y,
  X = X,
  family = binomial(),     # Use binomial for binary classification (screening)
  SL.library = sl_library,
  cvControl = list(V = 5), # 5-fold cross-validation
  method = "method.NNLS"   # Non-negative least squares to find optimal weights
)

# View the results
summary(cv_sl)
```

### Support Vector Machine Base Learner

```{r svm-cv}
# Create a custom SVM with a linear kernel
SL.ksvm.linear <- function(...) {
  SL.ksvm(..., kernel = "vanilladot") # 'vanilladot' is linear in kernlab
}

# Add it to library
sl_library <- c("SL.ksvm.linear")

# Run the Cross-Validated SuperLearner
set.seed(542)
cv_sl <- CV.SuperLearner(
  Y = Y,
  X = X,
  family = binomial(),
  SL.library = sl_library,
  cvControl = list(V = 5),
  method = "method.NNLS"
)

# Compare performance
summary(cv_sl)
plot(cv_sl) # Visual comparison of risk (MSE) for each learner
```

### XGBoost Base Learner

```{r xgb-cv}
# SL.xgboost default: ntrees = 1000, max_depth = 4, shrinkage = 0.1
sl_library <- c("SL.xgboost")

# Optional: Create custom XGBoost configurations (Hyperparameter Tuning)
SL.xgb.shallow <- function(...) SL.xgboost(..., max_depth = 2, ntrees = 500)
sl_library <- c(sl_library, "SL.xgb.shallow")

# Run the Cross-Validated SuperLearner
set.seed(542)
cv_sl <- CV.SuperLearner(
  Y = Y,
  X = X,
  family = binomial(),     # Necessary for binary full-text screening
  SL.library = sl_library,
  cvControl = list(V = 5), # 5-fold cross-validation
  method = "method.NNLS"
)

# Evaluate Results
summary(cv_sl)
```

### Naive Bayes Base Learner

*(Add Naive Bayes learner setup here as needed.)*

## SuperLearner Ensemble

### Convert data for the ensemble

```{r convert-data}
# Convert to data.table for speed
dt <- data
dt <- as.data.table(dt)
```

### 3. Separate outcome and predictors

```{r ensemble-outcome-predictors}
Y <- dt[[2]]
X <- dt[, -c(1, 2), with = FALSE]
```

### 4. Define SuperLearner base learners

```{r ensemble-learners}
learners <- c(
  "SL.glm",           # Logistic Regression
  "SL.svm",           # Support Vector Machine
  "SL.xgboost",       # XGBoost
  "SL.randomForest",  # Random Forest
  "SL.naiveBayes"     # Naive Bayes
)
```

### 5. Run 5-fold SuperLearner cross-validation

```{r ensemble-cv}
set.seed(542)

sl_cv <- CV.SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),     # binary screening decision
  V = 5,                   # 5-fold cross-validation
  method = "method.NNLS",  # non-negative least squares
  cvControl = list(stratifyCV = TRUE)
)
```

### 6. Inspect ensemble weights

```{r ensemble-weights}
# Higher weight => learner contributes more to screening decisions
# Often: logistic + xgboost dominate text-embedding ensembles
sl_cv$SL.weights
```

### 7. Evaluate screening performance

```{r roc-auc}
# A. ROC-AUC (standard)

library(pROC)

roc_obj <- roc(Y, sl_cv$SL.predict)
auc(roc_obj)
```

```{r recall-precision}
# B. Recall-focused evaluation (critical for screening)

threshold <- 0.2  # typical for high-recall screening

pred <- ifelse(sl_cv$SL.predict >= threshold, 1, 0)

recall <- sum(pred == 1 & Y == 1) / sum(Y == 1)
precision <- sum(pred == 1 & Y == 1) / sum(pred == 1)

recall
precision
# In systematic reviews, recall >= 0.95 is usually required
# Precision can be low if recall is high
```

### 8. Identify false negatives (highest-cost errors)

```{r false-negatives}
# These are missed relevant papers, often reviewed manually.
false_negatives <- which(Y == 1 & pred == 0)

length(false_negatives)
```

### 9. Train final SuperLearner on full data (deployment model)

```{r final-model}
# After evaluation only:
# Use final_sl only for future unseen articles, not performance claims.
final_sl <- SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),
  method = "method.NNLS"
)
```

### 10. Class-weighted SuperLearner (critical for imbalance)

```{r class-weights}
# Why: Systematic review screening is usually extremely imbalanced (e.g., 2-10% relevant).
# Without weights, SuperLearner will under-optimize recall.

# A. Compute inverse-frequency weights

pos_rate <- mean(Y == 1)
neg_rate <- mean(Y == 0)

obs_weights <- ifelse(
  Y == 1,
  1 / pos_rate,
  1 / neg_rate
)
```

```{r weighted-cv}
# B. Pass weights into SuperLearner
sl_cv_weighted <- CV.SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),
  V = 5,
  obsWeights = obs_weights,
  method = "method.NNLS",
  cvControl = list(stratifyCV = TRUE)
)
```

### 11. Precision-Recall AUC (better than ROC for screening)

```{r threshold-selection}
# Why: Screening workflows require guaranteed recall (e.g., >= 95%), not arbitrary thresholds.

# A. Find lowest threshold achieving target recall

target_recall <- 0.95
threshold_grid <- seq(0.01, 0.5, by = 0.005)

threshold_perf <- sapply(threshold_grid, function(t) {
  pred <- ifelse(sl_cv_weighted$SL.predict >= t, 1, 0)
  sum(pred == 1 & Y == 1) / sum(Y == 1)
})

best_threshold <- min(threshold_grid[threshold_perf >= target_recall])

best_threshold
```

```{r workload-reduction}
# B. Compute workload reduction at that threshold

pred_final <- ifelse(sl_cv_weighted$SL.predict >= best_threshold, 1, 0)

workload_reduction <- 1 - mean(pred_final == 1)
workload_reduction

# This number is directly reportable. Often the key result in screening studies.
```

### 12. Probability calibration (Platt scaling)

```{r calibration-fit}
# Why: SuperLearner probabilities can be miscalibrated, especially with tree-based models.
# Calibration improves threshold stability.

# A. Fit calibration model using out-of-fold predictions
calibration_model <- glm(
  Y ~ sl_cv_weighted$SL.predict,
  family = binomial()
)
```

```{r calibration-predict}
# B. Generate calibrated probabilities
prob_calibrated <- predict(
  calibration_model,
  newdata = data.frame(sl_cv_weighted.SL.predict = sl_cv_weighted$SL.predict),
  type = "response"
)
```

```{r calibration-threshold}
# C. Re-optimize threshold on calibrated probabilities
threshold_perf_cal <- sapply(threshold_grid, function(t) {
  pred <- ifelse(prob_calibrated >= t, 1, 0)
  sum(pred == 1 & Y == 1) / sum(Y == 1)
})

best_threshold_cal <- min(threshold_grid[threshold_perf_cal >= target_recall])

best_threshold_cal

# Calibration is strongly recommended when using probabilities operationally
# Improves generalization to new batches
```
