---
title: "SuperLearner TI-IDF SVD"
author: "Heather, Jarrett, Andras"
output: html_document
---

## SuperLearner Ensemble

### 1. Overview

Base learners used:

- Elastic-net logistic regression (very strong baseline)
- Linear SVM
- XGBoost (low depth)
- Random forest (optional but weaker here)
- Naive Bayes (usually weak after SVD -- often drop)

### 2. Install packages and load data

```{r install, message=FALSE, warning=FALSE}
#install.packages(c(
#  "SuperLearner",
#  "glmnet",
#  "e1071",
#  "xgboost",
#  "randomForest",
#  "data.table",
#  "PRROC",
#  "pROC"
#))

```

```{r libraries, message=FALSE, warning=FALSE}
# Load all required libraries after installation
library(tidyverse)
library(SuperLearner)
library(xgboost)
library(glmnet) # Explicitly load glmnet
library(e1071)  # For SL.svm (even if commented out in learners, good practice)
library(randomForest)
#library(naivebayes)
library(data.table)
library(PRROC)
library(pROC)
library(parallel)
library(doParallel)
library(here)

```

### 1. Load in the Prepared Text Features and CV Folds

First we load in the prepared features from the previous steps.

- TF-IDV SVD features
- Doc2Vec features
- PubMedBERT embeddings

We also load in the cross-validation folds created in the data preparation step. This is critical for preventing leakage and ensuring that the cross-validation is properly structured according to the original data splits.

```{r load-data}
# Load data from all text preparation methods
dt_tfidf <- readRDS(here('04_output', "tfidf_svd_features.rds"))
dt_doc2vec <- readRDS(here('04_output', 'doc2vec_features.rds' ))
dt_pubmed <- readRDS(here('04_output', 'pubmedbert_embeddings.rds'))

#load in the folds created in the data preparation step. 
cv_folds <- readRDS(here('04_output',"cv_folds.rds"))

```

### 2. Custom SVM learner with class weights for imbalance 

The default SVM in SuperLearner is a nu-SVM which can perform poorly on imbalanced data. 
We define a custom C-SVM with class weights to handle the imbalance in our dataset. 
This is crucial for improving recall, which is our primary concern in screening tasks. 

```{r custom_svm}
#use a c-svm instead of the defult nu-svm for better performance on imbalanced data. Also set class weights to handle imbalance
SL.svm.custom <- function(Y, X, newX, family, obsWeights, ...) {
  library(e1071)
  fit <- svm(
    x = X, 
    y = as.factor(Y),
    type = "C-classification",  # C-SVM instead of nu-SVM
    kernel = "radial",
    class.weights = c("0" = 1, "1" = sum(Y == 0) / sum(Y == 1)),  # handle imbalance
    probability = TRUE
  )
  pred <- predict(fit, newdata = newX, probability = TRUE)
  pred <- attr(pred, "probabilities")[, "1"]
  return(list(pred = pred, fit = fit))
}

SL.naivebayes.custom <- function(Y, X, newX, family, obsWeights, ...) {
  library(naivebayes)
  fit <- naive_bayes(x = X, y = as.factor(Y))
  pred <- predict(fit, newdata = newX, type = "prob")[, "1"]
  return(list(pred = pred, fit = fit))
}
```

### 3. SuperLearner pipeline function with parallel processing

```{r}

superlearner_parallel_pipeline <- function(.data, cv_folds, run_test_sample = FALSE, test_sample_n_cols = 25) {
    
  #stop the previous cluster just in case it is still running from a previous run of the notebook. 
  #This prevents errors about clusters already running and ensures a clean start for parallel processing. 
 if(exists("cl")) {    
  suppressWarnings(stopCluster(cl))   
 }
    
  # Separate the response from features
  Y <- .data[[2]]
  X <- .data[, -c(1, 2)]
  
  if(run_test_sample ==TRUE) {
    # For quick testing, use a smaller sample of columns and rows. This is not for final results but can speed up iterations during development.
    set.seed(542)
    sample_columns <- sample(ncol(X), min(test_sample_n_cols, ncol(X)))
   
    X <- X[, sample_columns]
  }
  
  # Ensure that the response is mumeric
  Y <- as.numeric(Y)
  
  # Compute class weights
  pos_rate <- mean(Y == 1)
  neg_rate <- mean(Y == 0)
  
  # Assign inverse class frequency weights to balance the classes
  obs_weights <- ifelse(Y == 1, 1/pos_rate, 1/neg_rate)
 
  #set the number of cores for parallel processing. We use one less than the total to avoid overloading the system.
cl <- makeCluster(detectCores() - 1) 
#set up parallel backend to use the cluster
registerDoParallel(cl) 
#set a seed on each worker to ensure reproducibility across parallel processes
clusterSetRNGStream(cl, iseed = 42) 
#send SuperLearner to the cluster workers so they can run the learners in parallel
clusterEvalQ(cl, {library(SuperLearner); library(naivebayes)}) 
#send all necessary objects to each cluster worker so they can run the learners in parallel without errors
#the envir = environment() argument ensures that the objects are taken from the current environment of the function
#which is important to avoid issues with object not found errors in the workers as the default is for clusterExport to look in the global environment which may not have the necessary objects when running inside a function.xpor
clusterExport(cl, list(
  "Y", "X", "learners", "obs_weights", "cv_folds"
), envir = environment())

sl_cv <- CV.SuperLearner(
  Y = Y,
  X = X,
  SL.library = learners,
  family = binomial(),
  obsWeights = obs_weights,
  method = "method.NNLS",
  cvControl = list(V = 5, validRows = cv_folds), # when passing the folds set V here to the number of folds expected
  parallel = cl
)

stopCluster(cl)
  
return(sl_cv)

}

```

### 4. Define the SuperLearner library

Next we set up the SuperLearner library with our chosen base learners (Including the custom SVM we just defined).

- Logistic regression with elastic net regularization (glmnet).
- XGBoost
- Random forest
- Naive Bayes
- SVM with class weights (custom)


```{r learners}
# Set learners' library
learners <- c(
  "SL.glmnet",        # Elastic-net logistic (BEST for TF-IDF)
 # "SL.svm",           # Linear SVM
  "SL.xgboost"       # Shallow boosted trees
  #"SL.randomForest",    # Optional, usually lower weight
 #"SL.ranger",
  #"SL.naivebayes.custom",     # Naive Bayes
  #"SL.svm.custom" # Custom SVM with class weights for imbalance 
# "SL.ksvm",
 #"SL.bayesglm"
)
```


### 5. Run 5-fold SuperLearner cross-validation on all three feature sets (Doc2Vec, TF-IDF SVD, PubMedBERT) in parallel

- Use run_test_sample = TRUE to test on a sample of n columns for faster iteration during development. Set to FALSE for final runs on all features.

```{r run-superlearner}
doc2vec_res <- superlearner_parallel_pipeline(dt_doc2vec, cv_folds, run_test_sample = TRUE, test_sample_n_cols = 10)

tfidf_res <- superlearner_parallel_pipeline(dt_tfidf, cv_folds, run_test_sample = TRUE)
pubmed_res <- superlearner_parallel_pipeline(dt_pubmed, cv_folds, run_test_sample = TRUE)

doc2vec_res$coef

```


